\documentclass[11pt,english]{article}
\usepackage{babel}
\usepackage[numbers]{natbib}
%
\input{header-draft}
%
%
\begin{document}
\title{First steps towards a formalization of Forcing
  \\ Revisions made to the paper}
\author{E.~Gunther \and  M.~Pagano \and P.~S\'anchez~Terraf
  %% fourth author was partially
  %% supported by CONICET, ANPCyT project PICT 2012-1823, and by
  %% SeCyT-UNC project 05/B284.}
}
\date{}
\maketitle

In first place we want to thank the referees for reading the paper very
carefully and for their detailed and constructive criticism. Below we
detail the main changes that have been made; also, a diff-PDF is
provided that shows all the changes in color.

We have spelled-check the manuscript, corrected punctuation, and took
notice of all the typos indicated in the report. 

The Introduction has many changes. A whole new section (``Related
work'') has been added, where the formalization of mathematics in
general and of forcing and set theory in particular are discussed.
Also, at the end of the subsection on Isabelle we describe the tools
available for implementation, and a comparison
between the two dialects available, the basic procedural style and the
declarative \emph{Isar}. At some point of the Introduction, we also
remark that our 
development follows the presentation by Kunen \cite{kunen2011set}. 

We added several comments on the implementation details, especially in
the section on forcing notions; this includes a description of the
complexity in number of lines and proof commands, and some hints on
the mechanics of the use Isabelle.

A thorough comparison between the available results on $\AC$ and
$\DC$, and the need for reimplementing some of those was included in
the subsection on Dependent Choices.

The section on ``Names'' includes a thorough discussion on recursive
definitions in Isabelle/ZF and two new applications. The first is the proof
that the generic extension $M[G]$ is a transitive set; and the second
is that the generic filter $G$ belongs to $M[G]$, a fundamental result
that was missing from the previous two versions.

We take this opportunity to answer a question by Reviewer 3 from the
first round of corrections. Obtaining the results of this paper per se
was not very difficult. But as this Reviewer points out, the setting
up for the project, planning a strategy, and even studying Paulson's
libraries%
%%%%%%%%%%%%%%%%%%%%
\footnote{Even today we find ourselves inadvertently reproving some
  variant of a lemma in Paulson's development on set theory.}
%%%%%%%%%%%%%%%%%%%%
took almost an year of work. With the hindsight of this planning
period, we would like to point out that most of the results stated in
the conclusion section involve some serious development; and it might
even result in a publication for each item. It is therefore not
feasible to include any of them in this revision.
%
%
\bibliographystyle{arXiv/mi-estilo-else}
\bibliography{citados}
\end{document}
%%% Local Variables: 
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% End: 
